---
title: "Descriptive Statistics"
format: html
author: "Global Health and Infectious Diseases Group - KCCR-KNUST, Ghana for ISARIC"
self-contained: true
standalone: true
warnings: false
font-size: '14px'
---

## Descriptive Statistics

Descriptive statistics provide a summary of the main features of a dataset. They help us understand the characteristics of our data through measures of central tendency, dispersion, and distribution. This guide demonstrates how to compute and visualize descriptive statistics using R.

:::: {.callout-important}
**Before You Analyze**: Descriptive statistics are the foundation of any data analysis. Spend time understanding your data before running inferential tests. Always check for missing values, outliers, and data quality issues. "Look at your data" is the golden rule of statistics.
::::

### BLOCK 0: Helper (pretty separators in the Console)

:::: {.callout-note}
**Console Output**: We'll use a helper function to create visual separators in the console output, making it easier to track which block of code is running.
::::

```{r}
sep <- function(title) {
  cat("\n\n==================== ", title, " ====================\n", sep = "")
}
```

### BLOCK 1: Load packages & dataset

:::: {.callout-tip}
**Why Packages?** In R, we use packages to extend the functionality of the base R language. The tidyverse is a collection of R packages designed for data science, including `dplyr`, `ggplot2`, and `tidyr`. The `gtsummary` package provides functions for creating publication-ready summary tables.
::::

The `nrow()` function counts the number of rows (observations), and `ncol()` counts the number of columns (variables) in a data frame.

```{r, warning=FALSE, error=FALSE, message=FALSE}
sep("BLOCK 1: Load packages & dataset")

library(gtsummary)
library(tidyverse)

df <- read.csv(
  "https://raw.githubusercontent.com/EstebanGarciaG/Lectures/main/data/data_synth.csv"
)

# Quick preview
cat("Rows x Cols:", nrow(df), "x", ncol(df), "\n")
cat("Some columns:", paste(head(names(df), 10), collapse = ", "), "...\n\n")
head(
  df[, c(
    "demog_age",
    "dates_admdate",
    "outco_date",
    "adsym_fever",
    "comor_hypertensi",
    "demog_sex"
  )],
  5
)
```

### BLOCK 2: Clean types & compute length of stay (LOS)

:::: {.callout-warning}
**Data Cleaning**: Variables may not be in the correct format for analysis. Ensure continuous variables are numeric, dates are properly formatted, and categorical variables are appropriately encoded. This prevents calculation errors later.
::::

```{r, warning=FALSE, error=FALSE, message=FALSE}

sep("BLOCK 2: Clean types & compute LOS")

# Convert age to numeric (some may be character)
df$demog_age <- suppressWarnings(as.numeric(df$demog_age))

# Convert date columns to Date objects
df$outco_date <- as.Date(df$outco_date, format = "%Y-%m-%d")
df$dates_admdate <- as.Date(df$dates_admdate, format = "%Y-%m-%d")

# Compute LOS (days)
df$outco_lengthofstay <- as.numeric(df$outco_date - df$dates_admdate)
lengthofstay <- df$outco_lengthofstay

# Show quick summaries
cat("\nSummary: age\n")
summary(df$demog_age)
cat("\nSummary: LOS (days)\n")
summary(lengthofstay)
cat("\nMissing LOS count:", sum(is.na(lengthofstay)), "\n")
```

### BLOCK 3: Simple visualisations (histograms + boxplot)

:::: {.callout-tip}
**Exploratory Data Analysis**: Visualizations help identify data distribution, outliers, and patterns. Histograms show the shape of distributions, while boxplots display central tendency and spread at a glance.
::::

:::: {.callout-warning}
**Outlier Detection**: Pay attention to extreme values in histograms and boxplots. Outliers may represent data entry errors, genuine rare events, or patients with unusual characteristics that warrant separate analysis or clinical review.
::::

```{r, warning=FALSE, error=FALSE, message=FALSE}
sep("BLOCK 3: Simple visualisations")

# 2 histograms (age & LOS)
op <- par(mfrow = c(2, 1))
hist(df$demog_age, breaks = 30, main = "Age distribution", xlab = "Age (years)")
hist(lengthofstay, breaks = 30, main = "Length of stay", xlab = "Days")
par(op) # reset layout

# Boxplot of LOS
boxplot(lengthofstay, ylab = "Length of stay (days)", main = "Boxplot of LOS")
```

### BLOCK 4: Bar plot — fever by sex (stacked %)

:::: {.callout-note}
**Bivariate Analysis**: Combining two variables helps identify associations. Here we visualize fever prevalence by sex using proportions rather than raw counts, making comparisons easier across groups of different sizes.
::::

:::: {.callout-important}
**Proportions vs. Counts**: When comparing categorical variables across groups of different sizes, always report proportions or percentages. Raw counts can be misleading if one group is much larger than the other. For example, 50 cases out of 100 (50%) is proportionally the same as 100 cases out of 200 (50%), but reporting only raw counts (50 vs. 100) might suggest a difference that isn't there.
::::

```{r, warning=FALSE, error=FALSE, message=FALSE}
sep("BLOCK 4: Fever by sex (stacked %)")

# Keep only rows with non-empty fever values
df_clean <- subset(df, adsym_fever != "", select = c(demog_sex, adsym_fever))

# Contingency table: sex x fever
tab <- table(df_clean$demog_sex, df_clean$adsym_fever)

# Convert to within-sex proportions (percent)
prop_tab <- round(100 * prop.table(tab, margin = 1), 1)
cat("\nProportions of fever by sex (%):\n")
prop_tab

# Stacked barplot of proportions
op <- par(mar = c(5, 4, 4, 6)) # room for legend on the right
bp <- barplot(
  t(prop_tab),
  beside = FALSE,
  col = c("lightblue", "tomato"),
  ylim = c(0, 100),
  ylab = "Proportion (%)",
  xlab = "Sex",
  main = "Fever by sex (stacked %)"
)

# Add % labels inside bars
lbl_mat <- t(prop_tab)
y_mid <- apply(lbl_mat, 2, cumsum) - 0.5 * lbl_mat
text(
  x = rep(bp, each = nrow(lbl_mat)),
  y = as.vector(y_mid),
  labels = paste0(as.vector(lbl_mat), "%")
)

legend(
  "topright",
  inset = c(-0.25, 0),
  legend = colnames(prop_tab), # fever categories
  fill = c("lightblue", "tomato"),
  xpd = TRUE,
  bty = "n"
)
par(op)
```

### BLOCK 5: Create cleaned dataset for descriptive table

:::: {.callout-tip}
**Data Subsetting**: Select only variables relevant to your analysis. This improves code clarity, reduces errors from unused data, and makes the dataset more manageable for summary table creation.
::::

:::: {.callout-warning}
**Complete Case Analysis**: Removing rows with any missing values can substantially reduce your sample size and may introduce bias if data are not missing at random (MCAR). Always report how many observations were excluded and compare characteristics of included vs. excluded patients to assess for potential selection bias.
::::

```{r, warning=FALSE, error=FALSE, message=FALSE}
sep("BLOCK 5: Create cleaned dataset for descriptive table")

# Keep only variables of interest
keep_vars <- c(
  "demog_age",
  "outco_lengthofstay",
  "adsym_fever",
  "comor_hypertensi",
  "demog_sex"
)
df_subset <- df[, keep_vars]

# Helper to convert text to TRUE/FALSE, keep NA otherwise
to_logical <- function(x) {
  ifelse(
    x %in% c("True", "true"),
    TRUE,
    ifelse(x %in% c("False", "false"), FALSE, NA)
  )
}

# Recode binary-like text
df_subset$adsym_fever <- to_logical(df_subset$adsym_fever)
df_subset$comor_hypertensi <- to_logical(df_subset$comor_hypertensi)

# Ensure numeric for continuous
df_subset$demog_age <- suppressWarnings(as.numeric(df_subset$demog_age))
df_subset$outco_lengthofstay <- suppressWarnings(as.numeric(
  df_subset$outco_lengthofstay
))

# Drop rows with missing values in key fields
df_subset <- subset(
  df_subset,
  !is.na(demog_age) &
    !is.na(outco_lengthofstay) &
    !is.na(adsym_fever) &
    !is.na(comor_hypertensi) &
    demog_sex != ""
)

# Tidy sex as factor
df_subset$demog_sex <- factor(df_subset$demog_sex)

# Preview and size
cat("\nCleaned dataset preview:\n")
head(df_subset, 6)
cat(
  "\nCleaned dataset dims (rows, cols):",
  dim(df_subset)[1],
  dim(df_subset)[2],
  "\n"
)
```

### BLOCK 6: Build a simple descriptive "Table 1"

:::: {.callout-note}
**Summary Statistics Format**: Continuous variables are typically reported as median (IQR) with sample size. Binary/categorical variables are reported as count and percentage. This format is standard in clinical research Table 1 presentations.
::::

:::: {.callout-important}
**Choosing Central Tendency**: Use **mean ± SD** when data are normally distributed (symmetric, no extreme outliers). Use **median (IQR)** when data are skewed or contain outliers. Always inspect visualizations before choosing—the median is more robust to outliers and is preferred for most clinical outcomes.
::::

```{r}
sep("BLOCK 6: Descriptive table (Table 1-style)")

cont_vars <- c("demog_age", "outco_lengthofstay")
bin_vars <- c("adsym_fever", "comor_hypertensi")

# Groups: overall + by sex
groups <- c("All", levels(df_subset$demog_sex))

descriptive_table <- data.frame(
  Variable = character(),
  Group = character(),
  Summary = character(),
  stringsAsFactors = FALSE
)

for (g in groups) {
  dat <- if (g == "All") df_subset else subset(df_subset, demog_sex == g)

  # Continuous: median (IQR) | n
  for (v in cont_vars) {
    x <- dat[[v]]
    x <- x[is.finite(x)]
    med <- median(x, na.rm = TRUE)
    q1 <- quantile(x, 0.25, na.rm = TRUE)
    q3 <- quantile(x, 0.75, na.rm = TRUE)
    n <- length(x)
    descriptive_table <- rbind(
      descriptive_table,
      data.frame(
        Variable = v,
        Group = g,
        Summary = sprintf("%.1f (%.1f–%.1f) | %d", med, q1, q3, n),
        stringsAsFactors = FALSE
      )
    )
  }

  # Binary: count (%) | n
  for (v in bin_vars) {
    x <- dat[[v]]
    x <- x[!is.na(x)]
    n_yes <- sum(x == TRUE)
    prop_yes <- 100 * mean(x == TRUE)
    n_total <- length(x)
    descriptive_table <- rbind(
      descriptive_table,
      data.frame(
        Variable = v,
        Group = g,
        Summary = sprintf("%d (%.1f%%) | %d", n_yes, prop_yes, n_total),
        stringsAsFactors = FALSE
      )
    )
  }
}

# Markers (~ for continuous, * for binary)
descriptive_table$Variable <- ifelse(
  descriptive_table$Variable %in% cont_vars,
  paste0(descriptive_table$Variable, " ~: Median (IQR) | n"),
  paste0(descriptive_table$Variable, " *: Count (%) | n")
)

# Reshape to wide format (Table 1 style)
table_wide <- reshape(
  descriptive_table,
  idvar = "Variable",
  timevar = "Group",
  direction = "wide"
)

# Clean column names and print
names(table_wide) <- sub("^Summary\\.", "", names(table_wide))
cat("\nDescriptive table (wide):\n")
print(table_wide, row.names = FALSE)
```

### Bonus: Publication-Ready Summary Table with gtsummary

:::: {.callout-tip}
**Publication Ready**: The `gtsummary` package creates beautifully formatted summary tables suitable for research papers. The `tbl_summary()` function allows customization of labels, statistics, and decimal formatting.
::::

:::: {.callout-note}
**Stratification by Sex**: Reporting descriptive statistics overall and stratified by demographic subgroups (e.g., sex) helps readers assess whether characteristics are balanced and allows readers to identify patterns or disparities. Always compare baseline characteristics across groups to check for potential confounding.
::::

First, let's create age categories to make the age variable more interpretable in the summary table:

```{r}
df_subset_cleaned <- df_subset %>%
  mutate(
    age_category = case_when(
      demog_age < 20 ~ "<20 year",
      demog_age >= 20 & demog_age < 30 ~ "20-29 years",
      demog_age >= 30 & demog_age < 40 ~ "30-39 years",
      demog_age >= 40 & demog_age < 50 ~ "40-49 years",
      demog_age >= 50 & demog_age < 60 ~ "50-59 years",
      demog_age >= 60 & demog_age < 70 ~ "60-69 years",
      demog_age >= 70 & demog_age < 80 ~ "70-79 years",
      demog_age >= 80 ~ "80+ years",
      TRUE ~ NA_character_
    )
  )

# Basic summary table with raw variable names
df_subset_cleaned %>%
  dplyr::select(
    demog_age,
    age_category,
    demog_sex,
    adsym_fever,
    comor_hypertensi
  ) %>%
  tbl_summary()
```

Now let's create a polished version with custom labels and formatting:

```{r}
df_subset_cleaned %>%
  dplyr::select(
    demog_age,
    age_category,
    demog_sex,
    adsym_fever,
    comor_hypertensi,
    outco_lengthofstay
  ) %>%
  tbl_summary(
    label = list(
      demog_age ~ "Age, years",
      age_category ~ "Age category",
      demog_sex ~ "Sex",
      adsym_fever ~ "Fever on admission",
      comor_hypertensi ~ "Hypertension",
      outco_lengthofstay ~ "Length of stay"
    ),
    digits = all_categorical() ~ c(0, 1)
  ) %>%
  bold_labels()
```











