---
title: "6_Survival Analysis_CREDO"
format: html
---


### Intro to Survival Analysis
Survival analysis is a set of statistical methods used to study the time until an event occurs. In medical research, the event is often death, but it could also be recovery, disease recurrence, or hospital discharge.

Unlike other statistical methods, survival analysis takes into account not only whether an event occurred, but also when (i.e., time) it occurred. This makes it particularly useful for studies where participants are followed over time.

For example, imagine a study follows 1000 patients after they are admitted to hospital. Each patient is followed for up to one year.
During the study:
- Some patients die during follow-up.
- Some patients are still alive when the study ends.
- Some patients move away or are lost to follow-up before the study finishes.

For every patient, we record:

- The length of time they were followed.
- Whether they died or were still alive at their last follow-up.

These data can be used to estimate the probability that they survived over time, or even compare survival between different groups. 
Let's load our data and see how to do this in R.

### Loading packages and importing data

```{r}
# --- Packages
library(survival)

# --- Load data (your base)
df <- read.csv("https://raw.githubusercontent.com/EstebanGarciaG/Lectures/main/data/data_synth.csv")

# Convert types
df$demog_age     <- suppressWarnings(as.numeric(df$demog_age))
df$outco_date    <- as.Date(df$outco_date,    format = "%Y-%m-%d")
df$dates_admdate <- as.Date(df$dates_admdate, format = "%Y-%m-%d")

# Length of stay (days) from admission to outcome date
df$outco_lengthofstay <- as.numeric(df$outco_date - df$dates_admdate)

# Binary outcome already set by you: 1 = Death, 0 = Discharged alive
df$outco_binary_outcome <- ifelse(df$outco_binary_outcome == "Death", 1L, 0L)

# --- Prepare analysis dataset
df_km <- subset(
  df,
  is.finite(outco_lengthofstay) & outco_lengthofstay >= 0 & !is.na(outco_binary_outcome)
)

# Time and event
df_km$time  <- df_km$outco_lengthofstay
df_km$event <- df_km$outco_binary_outcome

# (Optional) nudge zero-length times for plotting stability
df_km$time[df_km$time == 0] <- 0.5


```



### Kaplan-Meier Survival Curves

The Kaplan–Meier method is used to estimate the probability of surviving over time.

Rather than giving a single percentage, it produces a survival curve, showing how the proportion of surviving patients changes throughout the follow-up period.

NB: The curve starts at exactly 100% survival at the beginning of the study and steps downward whenever a death occurs.

A Kaplan–Meier curve can answer questions such as:

- What proportion of patients are alive after x number of days?
- How does survival change over one year?
- At what point do deaths occur most frequently?

```{r}
# --- OVERALL Kaplan–Meier
fit_all <- survfit(Surv(time, event) ~ 1, data = df_km)

plot(fit_all,
     xlab = "Days since admission",
     ylab = "In-hospital survival probability",
     main = "In-hospital Survival (Kaplan–Meier)",
     conf.int = TRUE,
     mark.time = TRUE)



# We can also use the ggsurvfit package for a more polished plot. 
library(ggsurvfit)

ggsurvfit(fit_all) +
  labs(title = "In-hospital Survival (Kaplan–Meier)",
       x = "Days since admission",
       y = "In-hospital survival probability") +
  add_confidence_interval()+
  theme_bw() +
  theme(panel.grid = element_blank())
```



# Log-rank tests

The log-rank test compares the survival experiences of two or more groups. It asks a simple question: "Are the differences between these survival curves greater than would be expected by chance?"

NB: The log-rank test compares overall survival patterns but does not estimate the size of the difference or account for other patient characteristics.

```{r}
# --- BY-SEX Kaplan–Meier (colored) + log-rank
# Keep clear Male/Female labels only

df_km$demog_sex <- trimws(df_km$demog_sex)
df_km_sex <- subset(df_km, demog_sex %in% c("Male", "Female"))

# Fix factor order (so colors map predictably)
df_km_sex$demog_sex <- factor(df_km_sex$demog_sex, levels = c("Male", "Female"))

fit_sex <- survfit(Surv(time, event) ~ demog_sex, data = df_km_sex)

cols <- c("Male" = "steelblue", "Female" = "tomato")

# NB: the plot and legend have to be run together to ensure the colors match the curves.

plot(fit_sex,
     xlab = "Days since admission",
     ylab = "In-hospital survival probability",
     main = "In-hospital Survival by Sex (Kaplan–Meier)",
     conf.int = FALSE,   # set TRUE if you want per-curve CIs
     lwd = 2,
     mark.time = TRUE,
     col = cols[levels(df_km_sex$demog_sex)])

legend("bottomleft",
       legend = names(cols),
       col = unname(cols),
       lwd = 2, bty = "n",
       title = "Sex")



# Log-rank test
# This test compares the survival distributions of the two groups (sex). 

logrank <- survdiff(Surv(time, event) ~ demog_sex, data = df_km_sex)
print(logrank)

```



### Bonus
We can still use the `ggsurvfit` package for a more polished plot.
```{r}
ggsurvfit(fit_sex) +
  labs(title = "In-hospital Survival by Sex (Kaplan–Meier)",
       x = "Days since admission",
       y = "In-hospital survival probability") +
  add_confidence_interval()+
  theme_bw() +
  scale_color_manual(values = cols) +
  theme(legend.title = element_text(size = 10),
        legend.text = element_text(size = 9), 
        legend.position = "right",
        panel.grid = element_blank())

```



### Intro to Cox Proportional Hazards Regression

Cox proportional hazards regression is used to investigate how one or more factors influence the risk of death over time.

Unlike the Kaplan–Meier method, the Cox model can examine several variables simultaneously. For instance, we may wish to determine whether survival is associated with age, sex, comorbidity, etc. while adjusting for the effects of the other variables.

The main result (estimate) from a Cox model is the hazard ratio (HR).
The HR compares the risk of death between groups. Typically:

HR = 1.0 indicates no difference in risk.
HR > 1.0 indicates a higher risk of death.
HR < 1.0 indicates a lower risk of death.

```{r}

cox_age_only <- coxph(Surv(time, event) ~ demog_age, data = df_km_sex)
summary(cox_age_only)


cox_sex_only <- coxph(Surv(time, event) ~ demog_sex, data = df_km_sex)
summary(cox_sex_only)

cox_hypertension_only <- coxph(Surv(time, event) ~ comor_hypertensi, data = df_km_sex)
summary(cox_hypertension_only)

```


### Now, let's do a multivariable Cox regression including age, sex, and hypertension

```{r}
multivar_model <- coxph(Surv(time, event) ~ demog_age + demog_sex + comor_hypertensi, data = df_km_sex)


summary(multivar_model)


# Alternatively, we can present this using a nicer table using the `gtsummary` package.
library(gtsummary)

  tbl_regression(
    multivar_model,
    exponentiate = TRUE, 
    label = list(
      demog_age ~ "Age (years)",
      demog_sex ~ "Sex", 
      comor_hypertensi ~ "Hypertension"
    )) %>%
  as_gt() %>%
  gt::tab_header(
    title = "Multivariable Cox Proportional Hazards Model"
  )

```






















