---
title: "Basic R Programming"
author: "Introduction to R for Clinical Research"
format: html
self-contained: true
standalone: true
warnings: false
---

## Basic R Programming: A Comprehensive Guide

This guide covers fundamental R programming concepts needed for data analysis and statistical computing in clinical research.

## Getting Started with R

### What is R?

R is a free, open-source statistical programming language widely used in data analysis, clinical research, and biostatistics. It excels at:
- Statistical analysis and modeling
- Data manipulation and visualization
- Reproducible research workflows
- Creating interactive applications

### Installing R and RStudio

:::: {.callout-tip}
**Quick Setup**: 
1. Download R from https://www.r-project.org/
2. Download RStudio from https://posit.co/download/rstudio-desktop/
3. Install R first, then RStudio
4. Open RStudio and you're ready to code!
::::

## Basic Concepts

### Variables and Assignment

The **assignment operator** `<-` stores values in a variable for later use:

```{r}
# Create variables
age <- 45
name <- "Alice"
height <- 1.72

# View the variable
age
```

:::: {.callout-note}
**Assignment Operator**: Use `<-` or `=` to assign values. Most R programmers prefer `<-` for consistency.
::::

### Data Types

R has several fundamental data types:

| Type | Example | Description |
|------|---------|-------------|
| **Numeric** | `3.14`, `42` | Decimal or whole numbers |
| **Integer** | `5L`, `10L` | Whole numbers (with L suffix) |
| **Character** | `"text"`, `"Alice"` | Text strings (in quotes) |
| **Logical** | `TRUE`, `FALSE` | Boolean values |
| **NULL** | `NULL` | Empty/undefined value |

```{r}
# Check data type with class()
class(3.14)
class("hello")
class(TRUE)
```

### Vectors

A **vector** is a collection of values of the same type. Create vectors with `c()`:

```{r}
# Numeric vector
ages <- c(25, 30, 45, 52, 38)

# Character vector
names <- c("Alice", "Bob", "Carol", "David", "Eve")

# Logical vector
is_healthy <- c(TRUE, FALSE, TRUE, TRUE, FALSE)

# View the vector
ages
```

:::: {.callout-note}
**Vector Indexing**: Access specific elements using square brackets:
- `ages[1]` — first element
- `ages[c(1, 3)]` — first and third elements
- `ages[-2]` — all except the second element
::::

```{r}
# Vector indexing examples
ages[1] # First element
ages[c(1, 3)] # First and third
ages[-2] # All except second
```

### Data Frames

A **data frame** is a table of data, similar to a spreadsheet. Each column is a vector:

```{r}
# Create a data frame
patients <- data.frame(
  name = c("Alice", "Bob", "Carol"),
  age = c(45, 52, 38),
  bp_systolic = c(120, 135, 118)
)

# View the data frame
patients

# Access specific column
patients$name

# Access specific cell
patients$age[2]
```

## Operators

### Arithmetic Operators

```{r}
# Basic math
5 + 3 # Addition
10 - 4 # Subtraction
6 * 7 # Multiplication
20 / 4 # Division
2^3 # Exponentiation (2 to the power of 3)
17 %% 5 # Modulo (remainder: 17 divided by 5 = 3 remainder 2)
```

### Comparison Operators

```{r}
# Comparisons return TRUE or FALSE
5 > 3 # Greater than
5 < 3 # Less than
5 == 5 # Equal to
5 != 3 # Not equal to
5 >= 5 # Greater than or equal
5 <= 10 # Less than or equal
```

### Logical Operators

```{r}
# Combine multiple conditions
TRUE & FALSE # AND: both must be TRUE
TRUE | FALSE # OR: at least one must be TRUE
!TRUE # NOT: reverses the logical value
```

## Control Flow

### If/Else Statements

Execute different code based on conditions:

```{r}
age <- 45

if (age >= 65) {
  paste("Senior patient")
} else if (age >= 18) {
  paste("Adult patient")
} else {
  paste("Pediatric patient")
}
```

### For Loops

Repeat code a specific number of times:

```{r}
# Loop through numbers
for (i in 1:5) {
  paste(paste("Iteration", i))
}

# Loop through a vector
names <- c("Alice", "Bob", "Carol")
for (name in names) {
  paste(paste("Hello,", name))
}
```

### While Loops

Repeat code while a condition is TRUE:

```{r}
count <- 1
while (count <= 3) {
  paste(paste("Count:", count))
  count <- count + 1
}
```

## Functions: A Comprehensive Guide

### What is a Function?

A **function** is a reusable block of code that performs a specific task. Instead of writing the same code repeatedly, you define it once and call it many times with different inputs.

**Benefits of functions**:
- **Reduce repetition** — Write once, use many times
- **Improve readability** — Named functions explain intent
- **Enable testing** — Easier to verify correctness
- **Facilitate maintenance** — Update code in one place

### Anatomy of a Function

```{r}
# Basic function structure
my_function <- function(argument1, argument2) {
  # Function body: code that does the work
  result <- argument1 + argument2

  # Return statement: specify what the function outputs
  return(result)
}

# Call the function
my_function(5, 3)
```

:::: {.callout-note}
**Function Components**:
- **Name**: `my_function` — what you call it
- **Arguments**: `argument1, argument2` — inputs to the function
- **Body**: code between `{` and `}` — does the work
- **Return value**: what the function outputs (last line or `return()`)
::::

### Simple Example: Calculate BMI

```{r}
# Function to calculate Body Mass Index
calculate_bmi <- function(weight_kg, height_m) {
  # BMI = weight / height^2
  bmi <- weight_kg / (height_m^2)
  return(bmi)
}

# Use the function
calculate_bmi(70, 1.75)
calculate_bmi(65, 1.68)
```

### Functions with Default Arguments

Set default values so arguments are optional:

```{r}
# Function with default arguments
greet <- function(name, greeting = "Hello") {
  paste(greeting, name)
}

# Call with all arguments
greet("Alice", "Hi")

# Call with default greeting
greet("Bob")
```

### Functions with Multiple Returns

Return multiple values using a list:

```{r}
# Function returning multiple values
patient_summary <- function(name, age, bp) {
  category <- if (age >= 65) "Senior" else "Adult"
  bp_status <- if (bp >= 140) "High" else "Normal"

  result <- list(
    name = name,
    category = category,
    bp_status = bp_status
  )
  return(result)
}

# Call the function
summary <- patient_summary("Alice", 72, 145)
summary$category
summary$bp_status
```

### Functions with Variable Arguments

Use `...` to accept any number of arguments:

```{r}
# Function accepting variable arguments
sum_all <- function(...) {
  numbers <- c(...)
  total <- sum(numbers)
  return(total)
}

# Call with different numbers of arguments
sum_all(5, 10)
sum_all(5, 10, 15, 20)
sum_all(2, 4, 6, 8, 10, 12)
```

:::: {.callout-tip}
**The `...` (ellipsis)** allows functions to accept any number of arguments. Inside the function, collect them with `c(...)` to convert to a vector.
::::

### Common Function Patterns in Clinical Research

#### Pattern 1: Data Validation

```{r}
# Function with input validation
calculate_dose <- function(weight_kg, drug_dose_per_kg) {
  # Validate inputs
  if (weight_kg <= 0) {
    stop("Weight must be positive")
  }
  if (drug_dose_per_kg < 0) {
    stop("Dose cannot be negative")
  }

  dose <- weight_kg * drug_dose_per_kg
  return(dose)
}

# Valid call
calculate_dose(70, 0.5)

# Invalid call (will produce an error message)
# calculate_dose(-70, 0.5)
```

:::: {.callout-warning}
**Error Handling**: Use `stop()` to halt function execution with an error message when invalid inputs are detected.
::::

#### Pattern 2: Summary Statistics

```{r}
# Function computing multiple statistics
summarize_lab_values <- function(values) {
  # Remove missing values
  clean_values <- na.omit(values)

  # Calculate statistics
  stats <- list(
    n = length(clean_values),
    mean = mean(clean_values),
    sd = sd(clean_values),
    min = min(clean_values),
    max = max(clean_values),
    median = median(clean_values)
  )

  return(stats)
}

# Use the function
labs <- c(120, 125, 118, 140, 135, NA, 122)
summarize_lab_values(labs)
```

#### Pattern 3: Conditional Processing

```{r}
# Function that behaves differently based on input
classify_result <- function(value, test_name = "generic") {
  # Different reference ranges for different tests
  if (test_name == "glucose") {
    if (value < 100) {
      return("Normal")
    }
    if (value < 126) {
      return("Impaired Fasting Glucose")
    }
    return("Diabetic")
  } else if (test_name == "bp_systolic") {
    if (value < 120) {
      return("Normal")
    }
    if (value < 140) {
      return("Elevated")
    }
    return("High")
  } else {
    return("Unknown test")
  }
}

# Use the function
classify_result(115, "glucose")
classify_result(145, "bp_systolic")
```

### Scope: Local vs. Global Variables

Variables created inside a function exist only within that function (**local scope**). Variables created outside (**global scope**) are accessible everywhere:

```{r}
# Global variable
patient_name <- "Alice"

# Function using global variable
show_info <- function() {
  # This works because patient_name is global
  local_age <- 45

  paste(patient_name, "is", local_age, "years old")
}

show_info()

# This works
patient_name

# This fails because local_age only exists inside the function
# local_age
```

:::: {.callout-note}
**Best Practice**: Avoid relying on global variables inside functions. Pass needed values as arguments instead. This makes functions:
- More reusable
- Easier to test
- Less prone to bugs
::::

### Anonymous Functions

Functions without names, often used with `apply()` functions:

```{r}
# Calculate BMI for multiple weight-height pairs
heights <- c(1.65, 1.72, 1.80, 1.58)
weights <- c(60, 75, 85, 55)

# Using an anonymous function with mapply()
bmis <- mapply(
  function(w, h) {
    w / (h^2)
  },
  weights,
  heights
)
bmis
```

### Built-in Functions You'll Use Often

| Function | Purpose | Example |
|----------|---------|---------|
| `mean()` | Average | `mean(c(10, 20, 30))` → 20 |
| `sd()` | Standard deviation | `sd(c(10, 20, 30))` |
| `sum()` | Sum values | `sum(c(10, 20, 30))` → 60 |
| `length()` | Count elements | `length(c(10, 20, 30))` → 3 |
| `max()` / `min()` | Maximum/minimum | `max(c(10, 20, 30))` → 30 |
| `paste()` | Combine text | `paste("Age:", 45)` |
| `c()` | Combine into vector | `c(1, 2, 3)` |
| `data.frame()` | Create table | `data.frame(name, age)` |
| `head()` | First n rows | `head(df, 6)` |
| `summary()` | Quick summary | `summary(df)` |

## Working with Data

### Reading Data

```{r}
# From CSV file
df <- read.csv("data.csv")

# From URL
df <- read.csv("https://example.com/data.csv")

# View first few rows
head(df)
```

### Subsetting Data

```{r}
# Create sample data frame
patients <- data.frame(
  name = c("Alice", "Bob", "Carol"),
  age = c(45, 52, 38),
  diagnosis = c("HTN", "DM", "HTN")
)

# Select rows where age > 40
patients[patients$age > 40, ]

# Select specific columns
patients[, c("name", "age")]

# Using $ to access columns
patients$name
```

### Applying Functions to Data

```{r}
# Apply a function to each element
values <- c(10, 20, 30)
sqrt(values) # Square root of each value

# Apply to each column of a data frame
apply(patients[, 2:3], 2, mean) # Mean of numeric columns
```

## Best Practices

:::: {.callout-tip}
**Function Writing Best Practices**:

1. **Use descriptive names** — `calculate_bmi()` not `calc()`
2. **Keep functions small** — One task per function
3. **Document your code** — Add comments explaining logic
4. **Validate inputs** — Check that arguments make sense
5. **Return consistent types** — Always return the same data type
6. **Use meaningful argument names** — `weight_kg` not `w`
7. **Provide defaults** — When sensible, make arguments optional
::::

## Putting It All Together: A Clinical Example

```{r}
# Comprehensive example: Patient risk assessment

# Function to calculate cardiovascular risk score
assess_cv_risk <- function(age, sbp, ldl, hdl, smoker = FALSE) {
  # Input validation
  if (age < 0 | age > 120) {
    stop("Invalid age")
  }
  if (sbp < 50 | sbp > 300) {
    stop("Invalid blood pressure")
  }

  # Calculate risk components
  age_score <- if (age >= 60) {
    4
  } else if (age >= 50) {
    2
  } else {
    0
  }
  bp_score <- if (sbp >= 140) {
    4
  } else if (sbp >= 130) {
    2
  } else {
    0
  }
  ldl_score <- if (ldl >= 160) {
    4
  } else if (ldl >= 130) {
    2
  } else {
    0
  }
  hdl_score <- if (hdl < 40) {
    3
  } else if (hdl < 50) {
    1
  } else {
    0
  }
  smoke_score <- if (smoker) 4 else 0

  # Total risk score
  total_score <- age_score + bp_score + ldl_score + hdl_score + smoke_score

  # Classify risk
  risk_level <- if (total_score >= 12) {
    "High"
  } else if (total_score >= 8) {
    "Moderate"
  } else {
    "Low"
  }

  # Return results as list
  result <- list(
    total_score = total_score,
    risk_level = risk_level,
    components = list(
      age = age_score,
      bp = bp_score,
      ldl = ldl_score,
      hdl = hdl_score,
      smoking = smoke_score
    )
  )

  return(result)
}

# Use the function
patient1 <- assess_cv_risk(
  age = 65,
  sbp = 145,
  ldl = 150,
  hdl = 38,
  smoker = TRUE
)

patient1$risk_level
patient1$total_score
```

## Key Takeaways

1. **Functions are fundamental** to good R programming
2. **Write functions to avoid repetition** and improve code organization
3. **Validate inputs** to catch errors early
4. **Use meaningful names** for functions and arguments
5. **Return consistent data types** for predictability
6. **Document your functions** so others (and future you) understand them

## Resources

- [R Official Documentation](https://www.r-project.org/help.html)
- [R for Data Science](https://r4ds.hadley.nz/)
- [Advanced R](https://adv-r.hadley.nz/) — deeper dive into R programming
